Modal Abstractions for Virtualizing Memory Addresses

### 2025-03-23 programlogic.tex
finished advisor pass, switched to author-year citations, now 23 pages with a little wiggle room

\paragraph{Accessing Virtual Addresses}
Figure \ref{fig:wpdamd} includes two  rules for accessing memory at an address stored in a register $r_a$. 
Ultimately, any memory access needs to ensure the relevant address translation would succeed,
which can be ensured by what we informally call the page-table-walk points-to collection
($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}).
\TirNameStyle{WriteToRegFromVirtMem} and \TirNameStyle{WriteToVirtMemFromReg}
each use a virtual-pointsto assertion ($\textsf{vaddr} \mapsto_{\textsf{v}} \textsf{v}$),
and are nearly-standard (assembly) separation logic rules for memory accesses.
However, because we split the physical resources for the page table walk from the
virtual points-to itself (in order to permit the physical page tables to be freely modified
as long as they preserve virtual-to-physical translations), the rule requires $\mathcal{I}\textsf{ASpace}$
for the current address space to be carried through.
The soundness proofs for these rules extract
% the token ($\sumwalkabs\vaddr\qfrac\paddr$) from the virtual points-to
the token ($\fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac}$) from the virtual points-to,
use that to extract the physical page-table-traversal points-to collection describing
the page table walk for the relevant address ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$)
from the invariant ($\mathcal{I}\mathsf{ASpace}$), prove that the page table walk succeeds
and that memory or registers are updated appropriately, before re-packing the invariant and virtual points-to resources.

@@

\paragraph{Updating \lstinline|cr3|} 
Unlike other rules, \TirNameStyle{WriteToRegCtlFromRegModal} updates the root address of the 
address-space determining the validity of resources, from $\rtv$ before the
\lstinline|mov| to $\textsf{rvs}$ afterwards. The global effects of this rule are reflected in moving
bare assertions under an other-space modality for $\rtv$, and moving the new address space's assertions out of
the corresponding modality.
Note that the \emph{global} aspect is important, as a na\"ive frame rule would be unsound for updates to \lstinline|cr3|,
because one could frame out assertions in one address space, switch address spaces, and bring those assertions from the \emph{old}
address space back into the \emph{new} address space, where they may not hold. Appendix \ref{sec:issues} gives more details.
% From that rule, with one step outside the \textsf{vProp} logic (so it is not quite a derived rule within
% our embedded logic), we can derive 
% \TirNameStyle{WriteToRegCtlFromRegModal},
% which is the (full) Hoare double version of the address space switch rule proposed in Section \ref{sec:issues}.

>>>>

\subsubsection{Accessing Virtual Addresses}
Figure \ref{fig:wpdamd} includes two  rules for accessing memory at an address stored in a register $r_a$. 
Ultimately, any memory access needs to ensure the relevant address translation would succeed,
which can be ensured by what we informally call the page-table-walk points-to collection
($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}).
\TirNameStyle{WriteToRegFromVirtMem} and \TirNameStyle{WriteToVirtMemFromReg}
each use a virtual-pointsto assertion ($\textsf{vaddr} \mapsto_{\textsf{v}} \textsf{v}$),
and are nearly-standard (assembly) separation logic rules for memory accesses.
However, because we split the physical resources for the page table walk from the
virtual points-to itself (in order to permit the physical page tables to be freely modified
as long as they preserve virtual-to-physical translations), the rule requires $\mathcal{I}\textsf{ASpace}$
for the current address space to be carried through.
The soundness proofs for these rules extract
% the token ($\sumwalkabs\vaddr\qfrac\paddr$) from the virtual points-to
the token ($\fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac}$) from the virtual points-to,
use that to extract the physical page-table-traversal points-to collection describing
the page table walk for the relevant address ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$)
from the invariant ($\mathcal{I}\mathsf{ASpace}$), prove that the page table walk succeeds
and that memory or registers are updated appropriately, before re-packing the invariant and virtual points-to resources.

@@

\subsubsection{Updating \lstinline|cr3|} 
Unlike other rules, \TirNameStyle{WriteToRegCtlFromRegModal} updates the root address of the 
address-space determining the validity of resources, from $\rtv$ before the
\lstinline|mov| to $\textsf{rvs}$ afterwards. The global effects of this rule are reflected in moving
bare assertions under an other-space modality for $\rtv$, and moving the new address space's assertions out of
the corresponding modality.

Note that the \emph{global} aspect is important, as a na\"ive frame rule would be unsound for updates to \lstinline|cr3|,
because one could frame out assertions in one address space, switch address spaces, and bring those assertions from the \emph{old}
address space back into the \emph{new} address space, where they may not hold. 
% Appendix \ref{sec:issues} gives more details.
It is often said that the frame rule (below) is one of the key pieces of separation logic.
\[
  \mbox{$\inferrule*[right=Frame]{
    \{P\}\;C\;\{Q\}
  }{
    \{P\ast R\}\;C\;\{Q\ast R\}
  }$}
  \qquad\textrm{(unsound with address space changes)}
\]
Such a rule is normally recoverable from Hoare doubles (see, e.g., \citet{Chlipala2011Bedrock,Chlipala2013Bedrock}).
However, in the presence of address space changes the traditional frame rule is unsound.
Consider:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule{\ldots }{
    \{\textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{\textsf{Post}\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast \textsf{Post}\}
  }
\end{mathpar}
In this (broken!) hypothetical example,
both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. 
The essential problem is that the frame rule is motivated by local reasoning about local updates, but
a switch of address space is a \emph{global} change that may invalidate information about virtual addresses.
Thus framing around arbitrary \lstinline|cr3| updates is unsound --- hence the \emph{global} nature of \textsc{WriteToRegCtlFromRegModal} ---
though a variant the ensures the same \lstinline|cr3| value is installed before and after the framing
can be recovered.
